Mirror, mirror on the wall. Who's the best match of them all?

### 2023-11-29 PNAS/index.Rnw
Fix lingering issues.

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
<<setup, include = F>>=
library(knitr)
opts_chunk$set(error=F, warning=F, message = F, dpi=300, echo = F)
@

@@

\dropcap{I}n forensic evidence a single conclusion is often based on  many comparisons, either implicitly or explicitly.
Multiple comparisons are a persistent problem when developing statistical methods for scientific problems \citep{benjaminiControllingFalseDiscovery1995}.
In forensic statistics, researchers have been focusing on building up research databases to develop algorithms for objective evidence comparisons.
\hh{In practice, these algorithms then allow searches across (vast) databases and report the best $N$ matches to a particular piece of evidence, raising the critically important problem of close non-matches }\citep{pcast}.

@@

This issue has been raised in the past with regard to DNA \citep{thompson2003} and fingerprints \citep{koehler2021}: one of the issues \citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid bombing case was that the large size of the IAFIS database used to locate similar prints made it possible to locate `unusually' similar non-matches.
\svp{Very close non-matches are a direct consequence of a large number of comparisons; they will occur more frequently as databases increase in size and algorithms become more powerful.}
% In this paper, we discuss that very close non-matches are far from an unusual phenomenon  but a direct consequence of a large number of comparisons, and simply reflect the power of the search algorithms.

@@

Algorithms make a large number of comparisons \svp{that are not obvious to the user}
%out of a user's sight.
The cross-correlation function, for example, \svp{was} \citep{vorburgerApplicationsCrosscorrelationFunctions2011} one of the first quantitative measures \svp{proposed} in response to the NAS report \citep{nas2009}, and is used in \svp{many} pattern searching algorithms to find the best alignment between two sources and also evaluate the overall similarity.
Finding the best \svp{alignment} requires implicitly sliding one source across the whole length (for one-dimensional sources, such as striations) or area (for two dimensional sources, such as impression marks) of the other source \svp{ while keeping} track of \svp{the value of a} similarity measure.
\svp{This mirrors the forensic examination process, as examiners visually}
rotate or align two pieces of evidence under comparison microscopes.
In order to avoid embarassing false accusations and a further erosion of the public's trust in science, we need to address the problem of multiple comparisons by controlling their effect on false discovery rates.
%An explicit search, in constrast, occurs when we conduct database searches.
%These searches often return the $N$ best matches from the database, but this requires making comparisons to all of the database entries which meet the search criteria.

@@

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts which can be compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope in a process shown in \Cref{fig:wire-blade}, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire; depending on the tool design, there can be between 2 and 4 cutting surfaces in contact with the substrate.

@@

In order to calculate the number of comparisons, we define $\ell$ to be the length of the blade cut, and $d$ to be the diameter of the wire. Without loss of generality, we will assume that the wire is fully covered with striations suitable for comparison; if this is not the case, it is equivalent to changing the value of $d$.
At minimum, there must be $\ell/d$ comparisons in order to assess whether one side of the blade cut matches one side of the wire; in this case, the comparisons are independent of each other (conditional on striation patterns) because no overlapping data is used.
Of course, this overlooks the possibility of partial matches and assumes perfect alignment between the diameter of the wire and the portion of the blade used to cut the wire, which is unrealistic.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{../fig/wire-blade-comparison.png}
\caption{A comparison between a wire and a blade cut requires sliding the wire along the entire blade cut length to determine the best match (or whether there is a match).}\label{fig:wire-blade}
\end{figure}

@@

Compounding this problem, open-set studies of striated comparisons (bullets, firing pin impressions) \citep{bajic2020, mattijssen2021,best2022} suggest the false positive rate for these comparisons is between 0.004545 \citep{best2022} and 0.07235 \citep{mattijssen2021}; pooling data from these three studies yields a false positive rate of \Sexpr{sprintf("%.2f", fpr)}. %, which is high relative to the consequences of a false identification of evidence from different sources.

If we begin by considering the multiple comparison problem relative to the $\ell/d$ independent comparisons between a blade cut and a wire, we can calculate the overall false positive rate for all comparisons using $r$, the false positive rate for one comparison, and simple probability. Let $F_n:=\{\text{at least one false positive in }n\text{ comparisons}\}$. Then,
\begin{align}
P(F_1) &= r\nonumber\\
P(\overline{F_1}) &= 1-r\nonumber\\
P(F_n) &= \left[1-r\right]^n\nonumber\\
P(\overline{F_n}) &= 1 - \left[P(\overline{F_1})\right]^n = 1 - \left[1 - r\right]^n.\label{eq:fpr-control}
\end{align}
Using this relationship between $F_n$ and the false positive rate $r$, we can determine what $r$ must be in order to limit $P(F_n)$ to an acceptable level by solving for $r$: $r = 1 - \left[1 - P(F_n)\right]^{1/n}$.

@@

If we look at this data differently, we can assess the probability that $N$ comparisons are made accurately (e.g. there are no coincidental matches or examiner errors).
It is clear from \Cref{fig:Accuracy} that with existing false positive error rate estimates between 0.07 and 0.0045, the accuracy of the aggregate comparison between a wire and a tool becomes questionable, as the number of candidate alignments that must be examined is too high.
Even the most innocuous example we could come up with (small blade, only 2 cutting surfaces, and a relatively large wire) involved a minimum of \Sexpr{2*ncomp_min} completely independent comparisons.
As a result, it seems improbable that a wire comparison made under current protocols (with corresponding error rates) would have an overall accuracy rate which would be reliable enough to be presented at trial.
When we then consider that examiners would make cuts at multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons even in the simple case we have presented here, the probability of a coincidental match increases

>>>>

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
<<setup, include = F>>=
library(knitr)
opts_chunk$set(error=F, warning=F, message = F, dpi=300, echo = F)
options(digits = 4)
@

@@

\dropcap{I}n forensic evidence a single conclusion is often based on  many comparisons, either implicitly or explicitly.
Multiple comparisons are a persistent problem when developing statistical methods for scientific problems \citep{benjaminiControllingFalseDiscovery1995}.
In forensic statistics, researchers have been focusing on building up research databases to develop algorithms for objective evidence comparisons.
In practice, these algorithms then allow searches across (vast) databases and report the best $N$ matches to a particular piece of evidence, raising the critically important problem of close non-matches\citep{pcast}.

@@

This issue has been raised in the past with regard to DNA \citep{thompson2003} and fingerprints \citep{koehler2021}: one of the issues \citep{fine2006} leading to the wrongful accusation of Brandon Mayfield in the 2004 Madrid bombing case was that the large size of the IAFIS database used to locate similar prints made it possible to locate `unusually' similar non-matches.
Very close non-matches are a direct consequence of a large number of comparisons; they will occur more frequently as databases increase in size and algorithms become more powerful.
% In this paper, we discuss that very close non-matches are far from an unusual phenomenon  but a direct consequence of a large number of comparisons, and simply reflect the power of the search algorithms.

@@

Algorithms make a large number of comparisons \svp{that are not obvious to the user}
%out of a user's sight.
The cross-correlation function\citep{vorburgerApplicationsCrosscorrelationFunctions2011}, for example, was one of the first quantitative measures proposed in response to the NAS report \citep{nas2009}, and is used in many pattern searching algorithms to find the best alignment between two sources and also evaluate the overall similarity.
Finding the best alignment requires implicitly sliding one source across the whole length (for one-dimensional sources, such as striations) or area (for two dimensional sources, such as impression marks) of the other source while keeping track of the value of a similarity measure.
This mirrors the forensic examination process, as examiners visually rotate or align two pieces of evidence under comparison microscopes.
In order to avoid embarrassing false accusations and a further erosion of the public's trust in science, we need to address the problem of multiple comparisons by controlling their effect on false discovery rates.
%An explicit search, in constrast, occurs when we conduct database searches.
%These searches often return the $N$ best matches from the database, but this requires making comparisons to all of the database entries which meet the search criteria.

@@

A forensics examiner tasked with determining whether a wire in evidence was cut by a specific tool will create one or more blade cuts which can be compared to the cut surface of the wire.
These cuts are made in a sheet of material similar to the wire's composition, and may be performed at multiple angles, as the angle of the tool to the substrate can affect which striations are recorded on the substrate surface.
The blade cuts will then be compared to the wire under a comparison microscope, though eventually, automatic comparison algorithms may also be validated for lab use.
Each side of each blade cut will be compared to each side of the wire; depending on the tool design, there can be between 2 and 4 cutting surfaces in contact with the substrate.

@@

In order to calculate the number of comparisons, we define $\ell$ to be the length of the blade cut, and $d$ to be the diameter of the wire. Without loss of generality, we will assume that the wire is fully covered with striations suitable for comparison; if this is not the case, it is equivalent to changing the value of $d$.
\svp{An illustration of the sliding comparison process is shown in \Cref{fig:wire-blade}.}
At minimum, there must be $\ell/d$ comparisons in order to assess whether one side of the blade cut matches one side of the wire; in this case, the comparisons are independent of each other (conditional on striation patterns) because no overlapping data is used.
Of course, this overlooks the possibility of partial matches and assumes perfect alignment between the diameter of the wire and the portion of the blade used to cut the wire, which is unrealistic.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{../fig/wire-blade-comparison-label.png}
\caption{A comparison between a wire and a blade cut requires sliding the wire along the entire blade cut length to determine the best match (or whether there is a match). \svp{Surfaces shown are rendered 2D topographical scans of a wire and blade cut taken with a confocal light microscope.}}\label{fig:wire-blade}
\end{figure}

@@

Compounding this problem, open-set studies of striated comparisons (bullets, firing pin impressions) \citep{bajic2020, mattijssen2021,best2022} suggest the false positive rate for these comparisons is between 0.0045 \citep{best2022} and 0.072 \citep{mattijssen2021}; pooling data from these three studies yields a false positive rate of \Sexpr{sprintf("%.2f", fpr)}. %, which is high relative to the consequences of a false identification of evidence from different sources.

If we begin by considering the multiple comparison problem relative to the $\ell/d$ independent comparisons between a blade cut and a wire, we can calculate the overall false positive rate for all comparisons using $r$, the false positive rate for one comparison, and simple probability. Let $F_n:=\{\text{at least one false positive in }n\text{ comparisons}\}$. Then, $P(F_1) = r$, $P(\overline{F_1}) = 1-r$, $P(F_n) = [1-r]^n$, and $P(\overline{F_n}) = 1 - \left[P(\overline{F_1})\right]^n$.
Using this relationship between $F_n$ and the false positive rate $r$, we can determine what $r$ must be in order to limit $P(F_n)$ to an acceptable level by solving for $r$: $r = 1 - \left[1 - P(F_n)\right]^{1/n}$.

@@

If we look at this data differently, we can assess the probability that $N$ comparisons are made accurately (e.g. there are no coincidental matches or examiner errors).
It is clear from \Cref{fig:Accuracy} that with existing false positive error rate estimates between 0.07 and 0.0045, the accuracy of the aggregate comparison between a wire and a tool becomes questionable, as the number of candidate alignments that must be examined is too high.
Even the most innocuous example we could come up with (small blade, only 2 cutting surfaces, and a relatively large wire) involved a minimum of \Sexpr{2*ncomp_min} completely independent comparisons.
As a result, it seems improbable that a wire comparison made under current protocols (with corresponding error rates) would have an overall accuracy rate which would be reliable enough to be presented at trial.
When we then consider that examiners would make cuts at multiple angles \citep{baikerToolmarkVariabilityQuality2015}, increasing the number of comparisons even in the simple case we have presented here, the probability of a coincidental match increases further still.
